---
title: "Statistical Inference - Coursera"
author: "Jens Hooge"
date: "21.9.2014"
output: html_document
---

Summary
---

This is the project for the statistical inference class. We will use simulation 
to explore inference and do some simple inferential data analysis on the 
ToothGrowth dataset. The project consists of two parts:

1. Simulation exercises.
2. Basic inferential data analysis.

```{r, echo=FALSE, message=FALSE}
library(reshape2)
library(ggplot2)
library(gridExtra)
library(dplyr)
```

Simulation Excercise
---

For the simulation of the mean distribution of exponential distributions, 
40 samples have been drawn from an exponential distribution with a rate
factor of 0.2 and the mean as well as the standard deviation has been computed. 
This was repeated between 1000 times, and the density of this mean 
distribution was visualized together with a QQ Norm plot of the z-transformed 
sample [Fig.: 1]. The function for this computation can be seen in the following 
code chunk.

```{r, echo=TRUE}
simulateExpMean <- function(n, nosim, lambda) {
    mu <- rep(0, nosim)
    for (i in 1:nosim) {
        mu[i] <- mean(rexp(n, lambda))
    }
    return(mu)
}
```

```{r, echo=FALSE, message=FALSE}
set.seed(42)
lambda <- .2
n <- 40
nosim <- 1000
simulations <- data.frame(nosim_100=simulateExpMean(n, 100, lambda),
                          nosim_1000=simulateExpMean(n, 1000, lambda),
                          nosim_10000=simulateExpMean(n, 10000, lambda),
                          nosim_100000=simulateExpMean(n, 100000, lambda))
```

```{r, echo=FALSE}
mu_hat <- colMeans(simulations)
var_hat <- sapply(simulations, var)
```

```{r, echo=TRUE}
sTest <- shapiro.test(sample(simulations$nosim_1000, size=5000))
```

A Shapiro Wilk Test confirmed the Null-Hypothesis [p=`r sTest$p.value`], that 
the sample was drawn from a normal distribution with a sample mean of 
$\hat\mu$=`r round(mu_hat[2],4)` and a $\hat\sigma^2$=`r round(var_hat[2],4)`, 
which almost equal the theoretical distribution parameters
$\mu$=$\lambda^{-1}$=`r lambda^-1` and 
$\sigma^2$=$\frac{1}{n-1}\cdot\lambda^{-2}$=`r (1/(n-1))*lambda^-2`, 
with $\lambda$=`r lambda` and n=`r n`.

Computation of the coverage for the 95% confidence interval.
```{r, echo=TRUE}
mu_hat[2] + c(-1,1) * 1.96 * sqrt(var_hat[2])/sqrt(n)
```

Coverage above 95% confidence could be achieved for the interval [4.738, 5.232].

Basic inferential data analysis
---

```{r, echo=TRUE, message=FALSE}
data(ToothGrowth)
```

We will use the ToothGrowth dataset and provide a basic summary of the data.

```{r, echo=TRUE}
vitaminC <- filter(ToothGrowth, as.character(supp) == "VC")
orangeJuice <- filter(ToothGrowth, as.character(supp) == "OJ")
```

```{r, echo=TRUE}
summary(vitaminC$len); summary(orangeJuice$len); table(ToothGrowth$supp, 
                                                       ToothGrowth$dose)
```

We can see that there is a difference in the mean teeth length and that the 
data is balanced for both conditions, where the guinee pigs received orange juice
(OC) or vitamin C (VC). Whether this difference in mean [Fig.: 2] is significant was explored using
a simple paired t-test.

```{r, echo=TRUE}
t_test <- t.test(vitaminC$len, orangeJuice$len, paired=TRUE)
```

With a p-value of **0.0025**, we reject the null hypothesis and claim that there is a signficiant difference in the mean length of teeth under the two conditions **orange juice** and **vitamin C**. The
95% confidence interval was shown to be **[-5.991341, -1.408659]**
It was still unclear however, whether the dosage, in either of the conditions
had an influence on the tooth growth. We ran two paired t-tests for dosages lesser
or equal to 0.5 and greater than 0.5.

```{r, echo=TRUE}
t_test <- t.test(filter(vitaminC, dose <= .5)$len, 
                 filter(orangeJuice, dose <= .5)$len, 
                 paired=TRUE)
```

```{r, echo=TRUE}
t_test <- t.test(filter(vitaminC, dose > .5)$len, 
                 filter(orangeJuice, dose > .5)$len, 
                 paired=TRUE)
```

Under the two conditions, dosages below 0.5 a significant difference **[p = 0.01547]**,
with a 95% confidence interval of **[-9.236542, -1.263458]** in tooth growth was
shown, as well as for dosages above 0.5 **[p = 0.05482]**, with a 95% confidence
interval of **[-5.91682146, 0.06682146]**.

Conclusion
---

It could be shown that, the difference in tooth growth under two conditions, was
significantly different and tooth growth for dosages of up to 1 mg orange juice
outperformed the effect on tooth growth of pure vitamin c of the same concentration.

Appendix
---

## Fig 1 - Parameter Estimation via Simulation
```{r fig 1, echo=FALSE, fig.width=8, fig.height=4}
p1 <- ggplot() +  
        geom_density(data=simulations, aes(x=nosim_1000, y=..density..)) +
        geom_vline(xintercept = mean(simulations$nosim_1000), color="grey") +
        geom_vline(xintercept = 1/lambda, color="black") + 
        geom_text(aes(x=mean(simulations$nosim_1000), 
                      label="estimated", 
                      y=0.3), 
                      colour="grey", angle=90, 
                      vjust = 1.2, 
                      text=element_text(size=8)) +
        geom_text(aes(x=mean(simulations$nosim_1000), 
                      label="expected", 
                      y=0.3), 
                      colour="black", angle=90, 
                      vjust = -1, 
                      text=element_text(size=8)) +
        xlab(expression(paste("hat(mu) ", '=',
                               mean(simulations$nosim_1000)))) +
        labs(x="mu", y="Density") + 
        theme_bw()

p2 <- ggplot(data=simulations, 
             aes(sample = scale(nosim_1000))) + 
             stat_qq(alpha=0.3) + 
             geom_abline(intercept=0, slope=1) +
             theme_bw()

grid.arrange(p1, p2,
             ncol=2) 
```

## Fig 2 - Influence of Vitamin C on Tooth Growth
```{r fig2, echo=FALSE, fig.width=8, fig.height=4}
ToothGrowth$supp <- as.factor(ToothGrowth$supp)
ToothGrowth$dose <- as.factor(ToothGrowth$dose)
levels(ToothGrowth$supp) <- c("Orange Juice", "VitaminC")
ggplot(ToothGrowth, aes(dose, len)) + 
    geom_boxplot() +
    geom_jitter(size=2, alpha=0.6) +
    facet_grid(.~supp) +
    theme_bw()
```