Practical Machine Learning - Prediction Assignment Writeup
========================================================
```{r}
library(caret)
library(party)
# library(doMC)
# registerDoMC(32)
library(doSNOW)
cl <- makeCluster(4, type = "SOCK")
registerDoSNOW(cl)
```

## Cleaning up the Data
Let's read the data and get some idea about its size.
```{r}
testData <- read.csv("../data//pml-testing.csv", stringsAsFactors=TRUE)
trainingData <- read.csv("../data//pml-training.csv", stringsAsFactors=TRUE)

dim(testData)
dim(trainingData)
```
Some of the columns are all NAs, because those might become a problem in the downstream analysis, they should be removed.
```{r}
removeNACols <- function(df) {
  return(df[, colSums(is.na(df)) != nrow(df)])
}

testData <- removeNACols(testData)
trainingData <- removeNACols(trainingData)

dim(testData)
dim(trainingData)
```

## Feature Selection
Instead of using the correlation based feature selection algorithm by Mark Hall, proposed by the authors, I used elastic net to choose the most important features. (TODO: Give a proper reason here)
```{r}
# grid <- expand.grid(.mtry = seq(1,5,5))
# trainedModels <- train(classe ~ ., data = trainingData,
#                        method = "cforest",
#                        controls = cforest_unbiased(ntree = 10))
```

## Preprocessing and Cross Validation
```{r}
ctrl <- trainControl(
  method = "repeatedcv",  # cross-validation method
  number = 10,            # number of folds
  repeats = 10,           # number of complete sets of folds
  allowParallel = TRUE)
```

## Training the Classifier
```{r}
set.seed(42)
X <- trainingData[1:ncol(trainingData)-1]
Y <- trainingData$classe
grid <- expand.grid(.lambda = seq(0, 0.2, length=20),
                     .alpha = 1)
ptm <- proc.time()
trainedModels <- train(X, Y,
                       method = "glmnet",
                       tuneGrid = grid,
                       trControl = ctrl,
                       preProc = c("center", "scale"),
                       na.action=na.omit)
t <- proc.time() - ptm
```


## Prediction on the Testset

```{r}
summary(cars)
```

You can also embed plots, for example:

```{r fig.width=7, fig.height=6}
plot(cars)
```

